\chameleon{}: Untangling Type Errors Through Interactive Visualization and Exploration

### 2022-12-14 sections/chameleon.tex
Updated bib

\begin{figure}
    \centering
    \includegraphics[width=\linewidth,trim=0mm 35mm 0mm 0mm]{images/add-pair-example.pdf}
    \caption{
    A type error displayed in Visual Studio Code\cite{microsoft_visual_nodate} and the Haskell Vscode extension\cite{haskell_haskell_nodate}.
The expression \texttt{addPair} is blamed for causing the type error. This may not match the programmers' intention. 
    }
    \label{fig:motivation-example}
\end{figure}
\subsubsection{\textbf{Traditional type errors show only limited location}}
Haack and Wells~\cite{haack_type_2004} noted that ``\textit{Identifying only one node or subtree of the program as the error location makes it difficult for programmers to understand type errors. To choose the correct place to fix a type error, the programmer must find all the other program points that participate in the error.}'' The type error fig \ref{fig:motivation-example} can be fixed in multiple locations. For instance  replacing \texttt{['0'..'9']} on line 1 with \texttt{[0..9]}, or replacing \texttt{fst x} and \texttt{snd x} on line 2 with \texttt {read (fst x)} and \texttt{read (snd  x)}. In the type error message, only the \texttt{addPair} expression on line 4 was blamed.  In this small example, the whole context is visible, but the problem is worse for the programmer in large programs where the lines of code contributing to the type error are far apart in the source code.

\subsubsection{\textbf{Traditional type errors are biased}}
A common form of bias happens when a type error is reported in one expression, but it can occur in multiple other expressions as well. In fig \ref{fig:motivation-example}, the error message arbitrarily focuses the users' attention on only \texttt{addPair}, while ignoring that the literals in the definition of \texttt{u} may be incorrect. %Technically, this results from the side effect of different unification orders of the internal type-checking technique and has no bearing on what the programmers think and expect. 
Another form of bias is that traditional type errors are often framed as conflicts between \texttt{Expected type} and \texttt{Actual type}. This framing is standard practice in most typed languages. However, what is the \texttt{expected} and what is \texttt{actual} are a side effect of different unification orders rather than the intention of the programmer. In both of these cases, the error message may lead programmers to falsely believe the validity of parts of code and wrongly accuse others.

@@

% We propose candidate expression cards to make it clear to the user that there are multiple ways to resolve a type conflict. By contrast,  standard compiler error messages such as those of GHC arbitrarily focus the users' attention on only one candidate expression, e.g.\ ``\textit{Couldn't match expected type ‘Char’ with actual type ‘Bool’ in expression x}" where x is the candidate expression. In practice, this candidate expression often does not match programmers' intention (Fig. \ref{fig:single-candidate}).  
A candidate expression is an expression that can be inferred to have two conflicting types. 
When a type error is detected, \chameleon{} provides a list of all candidate expressions, and programmers are free to choose the problem to resolve by clicking on one candidate expression card. In the example shown in Fig. \ref{fig:expression}, \texttt{x} and \texttt{y} are both candidate expressions. Fixing either type error will make both expressions well-typed.

@@

% \begin{lstlisting} [
%     language=Haskell, caption={
%     A simple program that is ill-typed. This program generates two constraints from line 1 and one constraint from line 2.
% }, label={listing:ifelse}
% ]
% if a then b else c
% a = "True"
% \end{lstlisting}
\begin{listing}[!ht]
\begin{minted}{haskell}
    if a then b else c
    a = "True"
\end{minted}
\vspace{-3mm}
\caption{A simple program that is ill-typed. It generates two constraints from line 1 and one constraint from line 2. }
\label{listing:1}
\end{listing}
For instance, for the program in the listing \ref{listing:1}, \chameleon{}  generates the following constraints and labels (in brackets) $T_a = Bool$ (if condition), $T_b = T_c$ (if branches), $T_a= String$  (definition). Clearly, as $T_a$ can not unify with both \textit{Bool} and \textit{String}, this program is not well typed. \chameleon{} can construct a human-readable explanation from the MUS. An example output for listing \ref{listing:1} can be: \texttt{a} has type \texttt{Bool} because \texttt{a} is the condition of an if statement; however, \texttt{a} has type \texttt{String} because \texttt{a} is defined as the string literal \texttt{"True"}. This explanation facilitates the deduction steps (Section \ref{sub:deduction-steps}).

>>>>

\begin{figure}
    \centering
    \includegraphics[width=\linewidth,trim=0mm 35mm 0mm 0mm]{images/add-pair-example.pdf}
    \caption{
    A type error displayed in Visual Studio Code\cite{microsoft_visual_nodate} and the Haskell Vscode extension\cite{haskell_haskell_nodate}.
The expression \texttt{addPair} is blamed for causing the type error. This may not match the programmers' intention. 
    }
    \label{fig:motivation-example}
\end{figure}
\subsubsection{\textbf{Traditional type errors show only limited location}}
Haack and Wells~\cite{haack_type_2004} noted that ``\textit{Identifying only one node or subtree of the program as the error location makes it difficult for programmers to understand type errors. To choose the correct place to fix a type error, the programmer must find all the other program points that participate in the error.}'' The type error in Fig.~\ref{fig:motivation-example} can be fixed in multiple locations. For instance  replacing \texttt{['0'..'9']} on line 1 with \texttt{[0..9]}, or replacing \texttt{fst x} and \texttt{snd x} on line 2 with \texttt {read (fst x)} and \texttt{read (snd  x)}. In the type error message, only the \texttt{addPair} expression on line 4 was blamed.  In this small example, the whole context is visible, but the problem is worse for the programmer in large programs where the lines of code contributing to the type error are far apart in the source code.

\subsubsection{\textbf{Traditional type errors are biased}}
A common form of bias happens when a type error is reported in one expression, but it can occur in multiple other expressions as well. In Fig.~\ref{fig:motivation-example}, the error message arbitrarily focuses the users' attention on only \texttt{addPair}, while ignoring that the literals in the definition of \texttt{u} may be incorrect. %Technically, this results from the side effect of different unification orders of the internal type-checking technique and has no bearing on what the programmers think and expect. 
Another form of bias is that traditional type errors are often framed as conflicts between \texttt{Expected type} and \texttt{Actual type}. This framing is standard practice in most typed languages. However, what is the \texttt{expected} and what is \texttt{actual} are a side effect of different unification orders rather than the intention of the programmer. In both of these cases, the error message may lead programmers to falsely believe the validity of parts of code and wrongly accuse others.

@@

% We propose candidate expression cards to make it clear to the user that there are multiple ways to resolve a type conflict. By contrast,  standard compiler error messages such as those of GHC arbitrarily focus the users' attention on only one candidate expression, e.g.\ ``\textit{Couldn't match expected type ‘Char’ with actual type ‘Bool’ in expression x}" where x is the candidate expression. In practice, this candidate expression often does not match programmers' intention (Fig. \ref{fig:single-candidate}).  
A candidate expression is an expression that can be inferred to have two conflicting types. 
When a type error is detected, \chameleon{} provides a list of all candidate expressions, and programmers are free to choose the problem to resolve by clicking on one candidate expression card. In the example shown in Fig. \ref{fig:expression}, \texttt{x} and \texttt{y} are both candidate expressions. Fixing either type error can make both expressions well-typed.

@@

% \begin{lstlisting} [
%     language=Haskell, caption={
%     A simple program that is ill-typed. This program generates two constraints from line 1 and one constraint from line 2.
% }, label={listing:ifelse}
% ]
% if a then b else c
% a = "True"
% \end{lstlisting}
\begin{listing}[!ht]
\begin{minted}{haskell}
    if a then b else c
    a = "True"
\end{minted}
\vspace{-3mm}
\caption{A simple program that is ill-typed. It generates two constraints from line 1 and one constraint from line 2. }
\label{listing:1}
\end{listing}
For instance, for the program in Listing~\ref{listing:1}, \chameleon{}  generates the following constraints and labels (in brackets) $T_a = Bool$ (if condition), $T_b = T_c$ (if branches), $T_a= String$  (definition). Clearly, as $T_a$ can not unify with both \textit{Bool} and \textit{String}, this program is not well typed. \chameleon{} can construct a human-readable explanation from the MUS. An example output for Listing~\ref{listing:1} can be: \texttt{a} has type \texttt{Bool} because \texttt{a} is the condition of an if statement; however, \texttt{a} has type \texttt{String} because \texttt{a} is defined as the string literal \texttt{"True"}. This explanation facilitates the deduction steps (Section \ref{sub:deduction-steps}).

### 2022-12-14 sections/conclusion.tex
Updated bib

We present \chameleon{}, a type debugging tool for the Haskell programming language. Its constraint-based type inference engine provides unbiased and comprehensive error location reporting. \chameleon{} also features an interactive debugging interface with three debugging modes, enabling programmers to interactively explore the stepwise explanations. Our studies evaluated the tool's design with programmers. We found that, particularly for more complex tasks, \chameleon{} helped programmers to fix type errors more quickly than traditional text-based error messages. Further, programmers actively using \chameleon{} interactive features are shown to fix type errors faster than simply reading the type error output.

>>>>

We present \chameleon{}, a type debugging tool for the Haskell programming language. Its constraint-based type inference engine provides unbiased and comprehensive error location reporting. 
% This sentence could go away:
%\chameleon{} also features an interactive debugging interface with three debugging modes, enabling programmers to interactively explore the stepwise explanations. 
Our studies evaluated the tool's design with programmers. We found that, particularly for more complex tasks, \chameleon{} helped programmers to fix type errors more quickly than traditional text-based error messages. Further, programmers actively using \chameleon{} interactive features are shown to fix type errors faster than simply reading the type error output.
\chameleon{} currently works with the Haskell language, but in future we plan to extend the typechecking system to work with other strongly typed languages, such as Rust or TypeScript.

### 2022-12-14 sections/discussion.tex
Updated bib

\subsection{Externalise Intermediate Typing Inforamtion}
We speculate another factor of the effectiveness \chameleon{} interactive debugging tools is they help programmers effectively chunk intermediate information. With the program showing in the listing \ref{listing:2}, \chameleon{} offers two candidate expressions: \texttt{f} can be typed as \texttt{Int -> Bool} or \texttt{Char -> Bool}; \texttt{z} can be typed as \texttt{Int} or \texttt{Char}. Although, in theory, these two statements are to the same effect, programmers are often required to compute the latter from the former or vice versa. And this computation may carry out multiple layers. Programmers have to remember all the intermediate types and their reasoning throughout such mental gymnastics. Assisted by candidate expression cards and deduction steps, this intermediate information is externalized on screen and can be retrieved anytime. A recent study on working memory \cite{crichton_role_2021} suggested this approach may provide a positive effect in helping programmers manage cognitive load and free up working-memory space for high-level thinking.

>>>>

\subsection{Externalise Intermediate Typing Inforamtion}
We speculate another factor of the effectiveness \chameleon{} interactive debugging tools is they help programmers effectively chunk intermediate information. With the program shown in Listing~\ref{listing:2}, \chameleon{} offers two candidate expressions: \texttt{f} can be typed as \texttt{Int -> Bool} or \texttt{Char -> Bool}; \texttt{z} can be typed as \texttt{Int} or \texttt{Char}. Although, in theory, these two statements have the same effect, programmers are often required to compute the latter from the former or vice versa. And this computation may carry out multiple layers. Programmers have to remember all the intermediate types and their reasoning throughout such mental gymnastics. Assisted by candidate expression cards and deduction steps, this intermediate information is externalized on screen and can be retrieved anytime. A recent study on working memory \cite{crichton_role_2021} suggested this approach may provide a positive effect in helping programmers manage cognitive load and free up working-memory space for high-level thinking.
